Manuscript_draft

### 2021-02-22 manuscript_draft.Rmd
work on manuscript

### Stand attributes by year
By year, mean plot volume, basal area and stand density did not differ significantly between both harvested units (i.e., p>0.05, `r appNums("standFitM.contrasts",display="cite")` and `r figNums("standPlots",display="cite")`).  Contrasts between the high-removal and control treatment units showed that volume, basal area and stand density differences were highest and significant (i.e., p<0.05) in the 1992 measurement year, and these differences diminshed over each subsequent measurement year.  In the 2019 measurement, volume and stand density were not significantly different between these two treatment units (i.e., p>0.05 for both).  Contrasts between the low-removal and the control treatment units showed a similar trend with basal area and stand density; the control unit had significanly higher basal area and stand density means in 1992, but the difference between the units diminished over each subsequent measurement, becoming non-significant (i.e., p>0.05) in 2019.  Plot means of volume did not differ significantly between the treatment units in any measurement year.

@@

In both species, post-treatment growth rates increased with the tree size at the time of treatment.  This has been found in other studies as well. and is likely explained by the assymetric competition theory. Implications of this result are that growth increased from treatment will be concentrated in the larger tree size classes.  If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.  Forest managers should account for this size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

@@

Volume

Diameter distributions
All units in this stand maintained a negative-exponential diameter distribution whereby the number of trees was highest in the smallest diameter classes.  Insofar as this diameter distribution indicates an uneven-aged stand structure, thinning in these stands maintained these characteristics and did not substantially alter the structure.  The species composition between the low removal and control units was also similar, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was leading the species in the larger classes.  This may reflect species differences in regeneration and growth.  In closed forests with undisturbed forest floors, Subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.  Spruce, on the other hand, can live for over 400 years and thus can achieve larger sizes and dominant canopy positions than fir.  This leads to a common stand structure in the Interior Region of BC that was found in the low-reomval and control untis.

The high-removal unit had a somewhat different species composition.  Here, fir was the leading sprecies in all diameter classes.  This means that the high-removal thinning treatment converted the stand compsition to fir-leading, which is a primary concern for many foresters in British Columbia.  Spruce is generally much preferred as a crop species over subalpine fir.  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and our results show that spruce increased its composition compared to fir over the 27-year period of this study.  In contrast, the ratio of spruce to fir volume decreased over the same period in both the low-removal and control units.  Over a timespan of several decades, it seems reasonable to anticipate that spruce:fir ratios will converge between all treatment units.

Stand mortality increased with basal area after treatment.  This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.  Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.

>>>>

### Stand attributes by year
By year, mean plot volume, basal area and stand density did not differ significantly between both harvested units (i.e., p>0.05, `r appNums("standFitM.contrasts",display="cite")` and `r figNums("standPlots",display="cite")`).  Contrasts between the high-removal and control treatment units showed that volume, basal area and stand density differences were highest and significant (i.e., p<0.05) in the 1992 measurement year, and these differences diminshed over each subsequent measurement year.  For example, mean plot density in the high-removal treatment unit was approximately 600 trees per hectare less than the control in 1994 (p<0.05); by 2019, that difference had decreased to 219 trees per hectare (p>0.05).  Contrasts between the low-removal and the control treatment units showed a similar trend with basal area and stand density; the control unit had significanly higher basal area and stand density means in 1992, but the difference between the units diminished over each subsequent measurement, becoming non-significant (i.e., p>0.05) in 2019.  Plot means of volume did not differ significantly between the treatment units in any measurement year.

@@

In both species, post-treatment growth rates increased with the tree size at the time of treatment.  This has been found in other studies as well. and may be explained by the asymmetric competition theory whereby larger trees exert more competition on smaller trees. Implications of this result are that growth increases from treatment will be predominantly concentrated in the larger tree size classes.  If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.  Forest managers should account for potential size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

@@

Stand density 
The high-removal treatment unit increased stand density more than the control unit over the 27-year study period and the differences between control and treatment units decreased in each subsequent measurement period.  In 2019, there were no significant stand density differences between all units.  These results suggest that the thinned stands increased density faster than the control and within three decades, there are no significant differences between the three units.  We note that these estimates are for trees with a minimum diameter of 7.5cm at breast height, and therefore our data do not account for new seedlings that may have become established over the time period of the study.  Thus, the phenomenon of increasing stand density may simply indicate that small trees responded with faster diameter growth in the thinned treatments (as previously discussed) and were able to achieve the minimum diameter class in this study. Increasing densities in the smaller diameter classes also suggests that these stands will maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.  Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.

The low-removal thinning treatment also maintained a species composition similar to the control, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.  In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Compared to spruce, subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.  Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.  This leads to a stand structure that is common in the Interior Region of BC and also found in the low-removal and control units.  The high-removal unit had a somewhat different species composition.  Here, fir was the leading species in all diameter classes.  This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and spruce increased its composition compared to fir over the 27-year period of this study.  Over time, we expect that spruce would continue to increase its proportion relative to fir, resulting in similar species composition - diameter distributions as the other units.

Volume and basal area
Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that increment in these variables did not differ much between the two.  Thus, the long-term implications of this is that these stands are on similar development trajectories

Over the 27-year period, the high-removal treatment unit accrued ca. 8m2 per hectare basal area more than the two other units.

Volume increment did not significantly differ between the units.

Stand mortality
Stand mortality increased with basal area after treatment.  This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.  Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.
